---
title: "Cyclistic Analysis"
output: html_document
date: "2023-09-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries
```{r cars}
library(tidyverse) #data wrangling
library(lubridate) #for date
library(ggplot2) #data viz
```

## Read CSV files
```{r pressure, echo=FALSE}

data0423 <- read.csv("202304-divvy-tripdata.csv")
data0423["month"] <- "April"

data0323 <- read.csv("202303-divvy-tripdata.csv")
data0323["month"] <- "March"

data0223 <- read.csv("202302-divvy-tripdata.csv")
data0223["month"] <- "February"

data0123 <- read.csv("202301-divvy-tripdata.csv")
data0123["month"] <- "January"

data1222 <- read.csv("202212-divvy-tripdata.csv")
data1222["month"] <- "December"

data1122 <- read.csv("202211-divvy-tripdata.csv")
data1122["month"] <- "November"

data1022 <- read.csv("202210-divvy-tripdata.csv")
data1022["month"] <- "October"

data0922 <- read.csv("202209-divvy-publictripdata.csv")
data0922["month"] <- "September"

data0822 <- read.csv("202208-divvy-tripdata.csv")
data0822["month"] <- "August"

data0722 <- read.csv("202207-divvy-tripdata.csv")
data0722["month"] <- "July"

data0622 <- read.csv("202206-divvy-tripdata.csv")
data0622["month"] <- "June"

data0522 <- read.csv("202205-divvy-tripdata.csv")
data0522["month"] <- "May"

```

## Check data structure
```{r}
# str(data0423)
# str(data0323)
# str(data0223)
# str(data0123)
# str(data1222)
# str(data1122)
# str(data1022)
# str(data0922)
# str(data0822)
# str(data0722)
# str(data0622)

```


##Combine dataset
```{r}
full_year_2223data <- rbind(data0423,data0323,data0223,data0123
                            ,data1222,data1122
                            ,data1022,data0922,data0822,data0722,
                            data0622,data0522)
```

##Take a peek of the dataset
```{r}
head(full_year_2223data)
str(full_year_2223data)
```

##Change started_at & ended_at to POSIXct objects
```{r}
full_year_2223data$started_at <- ymd_hms(full_year_2223data$started_at)
full_year_2223data$ended_at <- ymd_hms(full_year_2223data$ended_at)
```

###Extract the time & date components 
```{r}
full_year_2223data$time_start <- 
  format(full_year_2223data$started_at, format = "%H:%M:%S")
full_year_2223data$date_start <- 
  format(full_year_2223data$started_at, format = "%Y-%m-%d")
full_year_2223data$time_end <- 
  format(full_year_2223data$ended_at, format = "%H:%M:%S")
full_year_2223data$date_end <- 
  format(full_year_2223data$ended_at, format = "%Y-%m-%d")
```


##Convert time_start & time_end to POSIXct objects
```{r}
full_year_2223data$time_start <-
  as.POSIXct(full_year_2223data$time_start, format = "%H:%M:%S")
full_year_2223data$time_end <- 
  as.POSIXct(full_year_2223data$time_end, format = "%H:%M:%S")

class(full_year_2223data$time_start)
class(full_year_2223data$time_end)
```


##Obtain ride length 
```{r}
full_year_2223data$ride_length <- 
  difftime(full_year_2223data$time_end,full_year_2223data$time_start)%>%
  as.numeric( units = "mins")%>%
  round(digits=2)
```


##Create a Date object 
```{r}
full_year_2223data$date_start <- as.Date(full_year_2223data$date_start)
full_year_2223data$date_end <- as.Date(full_year_2223data$date_end)

class(full_year_2223data$date_start)
class(full_year_2223data$date_end)
```

##Obtain day of the week
```{r}
full_year_2223data$day_of_week <- weekdays(full_year_2223data$date_start)
```

#Remove any duplicates
```{r}
full_year_2223data$ride_id <- unique(full_year_2223data$ride_id)
```



#Descriptive statistics
```{r}
#average ride_length based on the group type 
aggregate(full_year_2223data$ride_length ~ 
            full_year_2223data$member_casual, FUN = mean)

#max ride_length based on the group type 
aggregate(full_year_2223data$ride_length ~ 
            full_year_2223data$member_casual, FUN = max)

#avg ride length based on the months
full_year_2223data$month_order <- factor(full_year_2223data$month, 
                                         levels = month.name)
aggregate(full_year_2223data$ride_length ~ 
            full_year_2223data$month_order, FUN = mean)

#avg ride length based on the day of the week
day_order <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
day_factor <- factor(full_year_2223data$day_of_week, levels = day_order)
result <- aggregate(full_year_2223data$ride_length ~ day_factor, FUN = mean)
print(result)
```


#Total Number of Rides by Day & Month
```{r}
#Total Number of Rides by Day (grouped by member/casual)
full_year_2223data %>%
  group_by(member_casual, day_of_week) %>%
  summarise(num_rides = n()) %>%
  arrange(member_casual, day_of_week) %>%
  mutate(day_of_week = factor(day_of_week, levels = c("Sunday", "Monday", "Tuesday", 
                                                     "Wednesday", "Thursday", "Friday", "Saturday"))) %>%
  ggplot(aes(x = day_of_week, y = num_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  xlab('Days of the Week') +
  ylab('Total Number of Rides') +
  ggtitle('Total Number of Rides by Day') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 



full_year_2223data %>%
  group_by(member_casual, month) %>%
  summarise(num_rides = n()) %>%
  arrange(member_casual, month) %>%
  mutate(month = factor(month, levels = month.name)) %>%
  ggplot(aes(x = month, y = num_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  xlab('Month') +
  ylab('Total Number of Rides') +
  ggtitle('Total Number of Rides by Month') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels by 45 degrees


```

#Day Statisitcs (Casual vs Member)
```{r}
#Total Number of Casual Rides by Day (group by rideable_type)
full_year_2223data %>%
  filter(member_casual == "casual") %>%
  group_by(rideable_type, day_of_week) %>%
  summarise(num_rides = n()) %>%
  arrange(rideable_type, day_of_week) %>%
  mutate(day_of_week = factor(day_of_week, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))) %>%
  ggplot(aes(x = day_of_week, y = num_rides, fill = rideable_type)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  xlab('Days of the Week') +
  ylab('Total Number of Rides') +
  ggtitle('Total Number of Casual Rides by Day') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels by 45 degrees

#Total Number of Member Rides by Day (group by rideable_type)
full_year_2223data%>%
  filter(member_casual == "member")%>%
  group_by(rideable_type,day_of_week)%>%
  summarise(num_rides=n())%>%
  arrange(rideable_type,day_of_week)%>%
  mutate(day_of_week = factor(day_of_week, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))) %>%
  ggplot(aes(x=day_of_week,y=num_rides,fill=rideable_type))+geom_col(position="dodge")+
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))+
  xlab('Days of the Week')+
  ylab('Total Number of Rides')+
  ggtitle('Total Number of Member Rides by Day')+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels by 45 degrees

```

#Month Statisitcs (Casual vs Member)
```{r}

#Total Number of Casual Rides by Month (group by rideable_type)
full_year_2223data%>%
  filter(member_casual == "casual")%>%
  group_by(rideable_type,month)%>%
  summarise(num_rides=n())%>%
  arrange(rideable_type,month)%>%
  mutate(month = factor(month, levels =month.name)) %>%
  ggplot(aes(x=month,y=num_rides,fill=rideable_type))+geom_col(position="dodge")+
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))+
  xlab('Month')+
  ylab('Total Number of Rides')+
  ggtitle('Total Number of Casual Rides by Month ')+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels by 45 degrees

#Total Number of Member Rides by Month (group by rideable_type)
full_year_2223data%>%
  filter(member_casual == "member")%>%
  group_by(rideable_type,month)%>%
  summarise(num_rides=n())%>%
  arrange(rideable_type,month)%>%
  mutate(month = factor(month, levels =month.name)) %>%
  ggplot(aes(x=month,y=num_rides,fill=rideable_type))+geom_col(position="dodge")+
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))+
  xlab('Month')+
  ylab('Total Number of Rides')+
  ggtitle('Total Number of Member Rides by Month ')+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels by 45 degrees



